Appendix A: Model Validation

### 2021-09-02 review/Response to reviews.Rmd

At the reviewer's suggestion, we have added a short paragraph to the introduction explaining why the pressures of habitat fragmentation and climate change might interact (i.e. be more than additive).

##  {#R1-3}

> One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
> As written in the methods and results, I found the implementation very difficult to follow.
> A slightly longer summary of the methods would help.

We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages (L229--239).
We are not entirely sure if this is the type of description the reviewer was looking for.
If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
See also our response to \@ref(R3-3).

@@

> In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
> While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit.
> One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn.
> Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).

Rather than comparing the fit of multiple models with different predictors, we took advantage of the `select = TRUE` option in the `gam()` function, which adds an additional penalty to smooth terms that effectively removes them from the model if they are not supported by data.
Therefore we cannot show something like a dAIC table.
However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term (Table 2).
We have replaced many of the parenthetical p-values reported in the results section with references to this table.
Additionally, we have added supplemental Figure S3 which shows the curves as separate panels for continuous forest and fragments with raw data superimposed.

@@

> In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.

Counter-intuitively, averaging weather conditions over an entire year may obscure ways plants perceive environmental variation.
For example, a drought in the wet season may have a different effect than a drought in the dry season.
This differential response to season- or month-level climate data is observable in yearly plant census.
Collapsing finer temporal scale environmental drivers to yearly averages (or a few means of arbitrarily chosen climate windows) is exactly what we avoid with this method.
The caveat to this is that without finer scale demographic data, one may not be able to discern the cause of an apparent delayed effect, which is why we expand on several possible mechanisms for delayed effects in detail in the discussion.

> While the authors mention this, what about adding something short about when using this type of analysis is critical and when it might not be (should all demographers adopt this approach? Why or why not?)

We have added additional detail to the conclusions recommending that all demographers consider lagged effects when possible in light of recent research (including this study) (L536--541).

@@

We agree with the reviewer that the exact timing of mortality within a season doesn't matter for population dynamics.
This is likely obvious to demographers who are used to working with yearly census data and finer scale climate data.
However, we think this paragraph is appropriate to follow up with the three before it---a delayed effect indicated by a DLNM plot like those in Figures 4, 5 and 6 could be due to 1) developmental delays, 2) buffering by below-ground storage tissues, 3) indirect effects, or 4) **none of the above.** We want to be sure to point out that in the case of survival especially, a "delayed effect" of SPEI 5 months in the past could simply mean that a plant died *during* a drought 5 months in the past (no actual delay between adverse conditions and death).
Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
We have revised this paragraph to emphasize the above points.

##  {#R2-21}

> The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
> acuminata.
> For instance, why does too much rain reduce survival?
> Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past? Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)?
> Etc.

While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now L465--469) and delayed effects on growth (L 445--450, now L475--480), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
We have added a paragraph to the discussion regarding effects of wet conditions on tropical plants.

@@

Some of the missing values in the data do represent *Heliconia* plants that lost all above-ground tissue but re-sprouted in successive years, but other missing values may simply be imperfect "recapture".

##  {#R3-2}

> Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames?
> You describe why not using longer smooths, but not shorter.

\
The choice of scale for the SPEI calculation is indeed difficult, and is related to the operational definition of drought (i.e. is drought defined as dry conditions over the previous 1, 2, 3 or more months?).
We are not certain that the reviewer's suggestion to allow the GAM to "do the smoothing" would have the same effect as changing the scale parameter when calculating SPEI.
SPEI is calculated using a sliding window, while GAMs use penalization to achieve smoothing.
We believe a thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper.
Additionally, the paper originally describing SPI recommends 3-month scale for droughts defined as depletion of soil moisture (McKee et al. 1993).

### 2021-09-08 review/Response to reviews.Rmd

We are delighted that the reviewer found the manuscript engaging and valuable.
Our responses to specific suggestions follow.

##  {#R1-1}

> By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
> The introduction could set up the reader better to understand both.
> Since drought is well covered, adding background and hypotheses for why extreme wet would have a negative effect on vital rates would help.

We edited a sentence in the introduction that cites Esteban et al. (2021) to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
*"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction..."* (L67--69).
Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon (L64).

@@

##  {#R1-2}

> In the intro and discussion, I think it would also be helpful to describe a little bit more about mechanisms for why extreme precipitation matters more in fragments compared to continuous forest.
> I wondered if the authors thought this was too obvious (edge effects, etc), but I think it'd be helpful to spell this out more clearly.
> L466 -- 467 offer a new hypothesis, but briefly revisiting the 'old' evidence/hypotheses also seems important.

At the reviewer's suggestion, we have added a short paragraph (starting on L74) to the introduction explaining why the pressures of habitat fragmentation and climate change might interact (i.e. be more than additive).

@@

We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs)), but showing the raw data made the plots visually cluttered, in our opinion.
Another way we approached this comment is by excluding areas of the heatmaps in figures 4-6 that are far from observed data, equivalent to not drawing regression lines beyond the range of data.
We implemented this with the `exclude.too.far()` function in the `mgcv` package.
Displaying raw data on the heatmaps was too messy and not informative in a 2d plot.

##  {#R1-5}

> I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.

The main limitations of this method are of data (see also reviewer 2's comment \@ref(R2-15)).
We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment (L542--550).

@@

> I had a hard time linking these statements in the results to the figures.
>
> L272--273---How do you see disparities in plant size here?
> I thought this was about growth?

Models which predict size in year t+1 as a function of size in year t, such as ours, are often described as growth functions because they can be used to project change in size (i.e. growth).
However, we agree that it would be more clear for this section heading to be "size" rather than "growth".

> L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
> More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
> More precise language would help.

We have replaced this sentence.
See our response to \@ref(R2-5).

@@

Our point was actually that SPEI had less of an impact on flowering probability in fragments.
We edited this sentence to read: "SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure 6b)."

##  {#R1-8}

> L442-443 -- What is a mechanism for why plants might flower more after drought?
> (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.

We are a little hesitant to speculate too much here, as not much is known about the optimal microclimate conditions for flowering in *H. acuminata*, but we have revised this paragraph (see also our response to \@ref(R2-17)).
It now includes the following possible explanation: *"Drought conditions could be favorable for* H. accuminata *flowering due to increased temperatures or decreased cloudiness associated with droughts (Pau et al. 2013),* *and the effects could be delayed due to the development time of inflorescences."*(L470-473).

@@

We are glad to hear the reviewer saw this work as important and well written.
We address specific concerns and suggestions below.

##  {#R2-1}

> The results are all about the effect of SPEI on plant vital rates, but the abstract implies that the manuscript presents the effect of temperature and precipitation on vital rates separately.

We have edited the abstract.
Now the sentence to which the reviewer refers reads: *"We used distributed lag non-linear models to understand the delayed effects of climate (measured as standardized precipitation evapotranspiration index, SPEI) on survival, growth, and flowering."* (L33).

@@

##  {#R2-2}

> Line 51-52.
> I find this sentence confusing.
> What specific demographic mechanisms are you referring to?
> Do you mean that high mortality rate lead to extinction and we don't know what causes high mortality rates?

The reference, Bruna et al. (2009), discusses the lack of knowledge of what demographic vital rates (e.g. survival, reproduction, recruitment) in which life stages (e.g. small or large individuals) contribute most to population declines in fragments.
We have re-ordered this sentence (now L50) to emphasize the hypothesis that fragmentation alters survival, growth, or reproduction with the caveat that few studies have determined which of these vital rate changes are driving extinctions.

@@

##  {#R2-3}

> Line 146.
> Can you explain what you mean with "the height of each plant to the tallest leaf"?
> If the tallest leaf is growing more horizontally the height of the plant would be shorter than the length of the largest leaf?

This is just the height of the plant from the ground to the tallest leaf leaf tip, not the tip of the youngest leaf.
The reviewer is correct that the height of a plant growing more horizontally (for example if it was knocked over by a branchfall) could be shorter than the length from the base to the end of the tallest leaf.
We edited the description to read: *"...the height of each plant (i.e. distance from the ground to the tallest leaf tip)"* (L152)

@@

##  {#R2-4}

> Lines 251ff: I don't see a reference of Fig 2 in the result section.

We have added a sentence to the results that refers to this figure (L284).

@@

##  {#R2-5}

> Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
> It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
> Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.

We thank the reviewer for pointing out this mistake.
We have edited this line to read "*Although plants were significantly larger in continuous forest compared to fragments, the effect of size in year t on size in year t+1 is nearly identical in the two habitats.*" (L303).
See also our response to \@ref(R1-figs).

@@

This sentence meant to describe the difference between CF and fragments, not the absolute size in fragments.
It has been re-worded to clarify (L348).

##  {#R2-10}

> Lines 355-359: The analysis did not consider testing a statistical difference between habitat type (see comment 7), so I think this sentence is misleading (overinterpreting the results).

While we could not test for an interaction between habitat and lagged SPEI (i.e. whether the shape of the crossbasis smooth was *significantly* different between habitats), we did test for a main effect of habitat by looking for overlap in 84% confidence intervals of model intercepts (see also response to \@ref(R2-6)).
However, to reduce confusion here, we have removed "significantly" so as not to imply statistical significance for all of "magnitude, directions and lag times" (L. 377).
See also our response to \@ref(R2-7).

@@

We have made this correction.

##  {#R2-17}

> Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?

We intended this paragraph to focus on one possible explanation for delayed effects---delays due to development times.
We believe that the bet-hedging argument was confusing for two reasons: 1) it assumed that drought (as we measured it with SPEI) is detrimental to *H. accuminata* survival, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
We have therefore revised the example in this paragraph (starting on L462) to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
See also our response to \@ref(R1-8).

@@

We agree with the reviewer and have replaced this sentence with a more in-depth explanation of Bruna et al. 2002 (L480).

##  {#R2-20}

> Lines 468 ff: I suggest deleting this paragraph.
> It doesn't really matter when exactly a plant dies during a growing season, all that matters for population dynamics is how many plants die from one growing season to the next.
> Distributed Lag Non-Linear Models account for the exact timing of the weather events.

We agree with the reviewer that the exact timing of mortality within a season doesn't matter for population dynamics.
This is likely obvious to demographers who are used to working with yearly census data and finer scale climate data.
However, we think this paragraph is appropriate to follow up with the three before it---a delayed effect indicated by a DLNM plot like those in Figures 4, 5 and 6 could be due to 1) developmental delays, 2) buffering by below-ground storage tissues, 3) indirect effects, or 4) **none of the above.** We want to be sure to point out that in the case of survival especially, a "delayed effect" of SPEI 5 months in the past could simply mean that a plant died *during* a drought 5 months in the past (no actual delay between adverse conditions and death).
Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
We have revised this paragraph (starting on L498) to emphasize the above points.

@@

##  {#R2-21}

> The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
> acuminata.
> For instance, why does too much rain reduce survival?
> Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past? Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)?
> Etc.

While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now L465--469) and delayed effects on growth (L 445--450, now L475--480), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
We have added a paragraph (L542) to the discussion regarding effects of wet conditions on tropical plants.

@@

Although the differences in plant size and flowering are relatively small, they are statistically significant (see Table 1 in revised manuscript), and we believe they are demographically significant over long time periods.
See our response to \@ref(R2-7).

##  {#R2-23}

> Line 499: This is confusing.
> If you measure size every year than you calculate growth as the change in size from one year to the next, don't you?

That's correct.
This sentence meant to point out that size is usually measured in yearly censuses while weather data can be more easily collected at monthly or daily timescales.
We've edited this sentence to read: "*The ability to identify size-specific lagged responses may be especially complicated given size is rarely measured at the same time scale (e.g. monthly) as climate drivers.*" (L533).

@@

```{r eval=FALSE, include=FALSE}
n_miss <- 
  model_data %>%
  group_by(year) %>%
  summarize(n_obs = n(), n_miss = sum(is.na(shts)), p_miss = n_miss/n_obs) %>% 
  slice(-1) %>%  #remove first year when all plants are found
  pull(p_miss)
1 - range(n_miss)

yrs_missing <- 
  model_data %>% 
  group_by(ha_id_number) %>% 
  filter(!is.na(last(shts))) %>% 
  summarize(n_obs = n(), n_miss = sum(is.na(shts))) %>% 
  filter(n_miss > 0) %>% 
  pull(n_miss)

quantile(yrs_missing, c(0.25, 0.5, 0.75, 0.95, 0.99))
```

We have edited to clarify: *"After the initial census year, between 80% and 97% of marked plants were found. Of plants that had missing values for some years, but were found again in a subsequent year, 95% had 2 or fewer years of missing values. Therefore, plants that were not found for three consecutive surveys, and no subsequent survey, were considered to have died in the transition year after their last observation."* (L159--164).

@@

\
The choice of scale for the SPEI calculation is indeed difficult, and is related to the operational definition of drought (i.e. is drought defined as dry conditions over the previous 1, 2, 3 or more months?).
We are not certain that the reviewer's suggestion to allow the GAM to "do the smoothing" would have the same effect as changing the scale parameter when calculating SPEI.
SPEI is calculated using a sliding window, while GAMs use penalization to achieve smoothing.
We believe a thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper.
Additionally, the paper originally describing SPI recommends 3-month scale for droughts defined as depletion of soil moisture (McKee et al. 1993).

##  {#R3-3}

> Line 191 - 194 (equation and description) for clarity you should describe what the index variables i and j are indexing, and I think clearer to have y_ij for ith plant in jth plot.
> Indexes on the random effect description suggest there is a separate variance estimated for each plot?
> d_it suggests that each plant has a separate SPEI estimate?
> I think it is probably aggregated at a higher level than individual?

We thank the reviewer for the very helpful comments on our model equation.
We have changed $y_i$ to $y_{ij}$ and added an explanation of the indexing as suggested.
We have also corrected the error in the random effect equation so that only one variance is included.
$d_{it}$ is intended to indicate that each plant observation is associated with a measure of SPEI for the past 36 months.
The SPEI history varies among plants spatially (there are three grid cells in the weather data within BDFFP, so different plots experience different climate) and temporally (observations from all years are pooled).

@@

This comment was especially helpful and made us re-think limiting the number of "knots" to 3 in the SPEI dimension.
The term *k* is probably more accurately described as the "maximum basis complexity" or "maximum basis dimension".
We had thought that by limiting k to 3, we were simply limiting the complexity of the smooth in the SPEI dimension.
However, after taking the reviewer's suggestions we realized that there were other unexpected consequences.
We increased *k* for the SPEI dimension for all models and for other smooths for some models.
This resulted in some changes in the the shape of the crossbasis smooth and which areas of it were statistically significant.
We have updated the methods and results sections accordingly.\
\
We have also now included a supplemental document with model validation, including checks that *k* is adequately large
.

##  {#R3-6}

> Line 211 - random effect -- how did you deal with this in the predictions?
> set to zero in design matrix, Choose a particular value, sample from the posterior?
> The code would tell me but not available.

What we show in the figures are not predictions (i.e. results of `predict()` in R), but partial effects plots generated with the `gratia` package (equivalent to `mgcv::plot.gam()`).
For figures 4, 5, and 6 we have added the model intercepts to evaluated smooths and back-transformed (when applicable) the estimates.
These plots, therefore, do not include the additive effects of other smooths, parametric covariates, or random effects.
We think that this is an intuitive way to display the results.
We added a paragraph to the methods (L260) explaining our strategy for visualizing the GAM results.

### 2021-09-08 doc/supplemental.Rmd
added reference to model validation appendix

---
title: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
author: 
  - Eric R. Scott
  - María Uriarte
  - Emilio M. Bruna
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
nocite: |
  @mcphadenChildProdigy1997981999, marengoDroughtAmazonia20052008, zengCausesImpacts20052008, lewis2010AmazonDrought2011
---

>>>>

---
title: "Appendix B: Supplemental figures and tables"
subtitle: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
author: 
  - Eric R. Scott
  - María Uriarte
  - Emilio M. Bruna
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
nocite: |
  @mcphadenChildProdigy1997981999, marengoDroughtAmazonia20052008, zengCausesImpacts20052008, lewis2010AmazonDrought2011
---

### 2021-09-08 doc/validate_models.Rmd
added figure to supplemental
added reference to model validation appendix

---
title: "Model validation plots"
author: "Eric R. Scott"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

@@

```{r eval=TRUE}
check_res_edf(s_1ha)
```
Adequate k for crossbasis.

@@

The crossbasis is much smoother

>>>>

---
title: "Appendix A: Model Validation"
subtitle: "Supplement to 'Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments'"
author: 
  - Eric R. Scott
  - María Uriarte
  - Emilio M. Bruna
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
linkcolor: blue
---

@@

```{r eval=TRUE}
# Function sourced from R/utils.R
check_res_edf(s_1ha)
```

@@

The crossbasis is smoother, but shows a similar pattern.
